### 2023-12-22 thesis/MSECE_thesis.tex
Last thesis edits.

The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep
learning-based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the
language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the
results in Section \ref{sec:pretraining_results} show that language models trained ``from scratch'' on in-domain data perform better than models
trained on out-of-domain data and transfer learned on in-domain data.

@@

\begin{equation}\label{eq:softmax}
    \sigma(x) = \frac{\exp(x_i)}{\sum_{j=1}^{N}\exp(x_j)} \quad \forall i=1..N
\end{equation}

@@

\begin{equation}\label{eq:scaled_dot_attention}
    \mbox{Attention}(Q,K,V) = \sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)V
\end{equation}

@@

Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP)
tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in
aviation English for domain-specific NLP tasks. However, the literature shows transformer-based language models have seen sparse use. The most notable
example of using a transformer neural network is the application of BERT for knowledge extraction from notice to airmen (NOTAM) messages
\cite{arnold_knowledge_2022}.

@@

\textbf{Different forms of English}. The BERT and RoBERTa pretraining checkpoints were created by training the models on literary works and typed
articles such as those from \textit{Wikipedia}. The data used in this work are orthographic transcriptions of aviation English, and while the two data
sources share the common root of English, transfer learning from literary English to aviation English yields limited results, as evidenced above,
especially with relatively small amounts of data.

@@

\subsection{Next Sentence Prediction}
This training procedure was introduced by Devlin et al.~\cite{devlin_bert_2019} as a pretraining procedure for BERT and is described in more detail in
Section \ref{sec:bert}. This can be used on top of masked language modeling for language model pretraining. However, it requires a knowledge of which
sequences in the corpus appear next to each other. For pretraining transformer-based neural language models for the aviation domain (with orthographic
transcriptions), this same procedure can be used to classify whether two transmissions occur in the correct order. In other words, it is classifying a
pilot's response to an air traffic controller directive or an air traffic controller's response to a pilot query. Of the four corpora used in this
work, ATCC is the only corpus that can be mined for an NSP-like pretraining procedure since the raw transcripts have indicators for time (start and
end indicators), transmitting party, and intended receiver. With this information, we can estimate the correct order of transmissions between two
parties to a reasonable degree of certainty (manually or programmatically).

>>>>

The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep
learning-based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the
language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the
results in Section \ref{sec:results_comparison} show that language models trained ``from scratch'' on in-domain data perform better than models
trained on out-of-domain data and transfer learned on in-domain data.

@@

\begin{equation}\label{eq:softmax}
    \mbox{softmax}(x) = \frac{\exp(x_i)}{\sum_{j=1}^{N}\exp(x_j)} \quad \forall i=1..N
\end{equation}

@@

\begin{equation}\label{eq:scaled_dot_attention}
    \mbox{Attention}(Q,K,V) = \mbox{softmax}\left(\frac{QK^T}{\sqrt{d_k}}\right)V
\end{equation}

@@

Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP)
tasks, more generally, in the conversational and written English domains, it follows that transformer architectures should see widespread use in
aviation English for domain-specific NLP tasks. However, the literature shows transformer-based language models have seen sparse use. The most notable
example of using a transformer neural network is the application of BERT for knowledge extraction from notice to airmen (NOTAM) messages
\cite{arnold_knowledge_2022}. While both BERT and RoBERTa have demonstrated that they perform well on English in general
\cite{devlin_bert_2019,liu_roberta_2019}, it is expected that the pretrained models will have limited performance due to the highly technical and
domain-specific nature of aviation English as well as the differing forms of the language; transcribed English speech as opposed to written English.
We expect the domain-specific terms and pronunciations will lead to a sufficiently different vocabulary from that of written English such that the
models trained from scratch will achieve better performance than the pretrained models.

@@

\textbf{Different forms of English}. The BERT and RoBERTa pretraining checkpoints were created by training the models on literary works and typed
articles such as those from \textit{Wikipedia}. The data used in this work are orthographic transcriptions of aviation English, and while the two data
sources share the common root of English, transfer learning from the casual style of written English to the technical, highly specific use of aviation
English yields limited results, as evidenced above, especially with relatively small amounts of data.

@@

\subsection{Next Sentence Prediction}
This training procedure was introduced by Devlin et al.~\cite{devlin_bert_2019} as a pretraining procedure for BERT and is described in more detail in
Section \ref{sec:bert}. This can be used on top of masked language modeling for language model pretraining. However, it requires a knowledge of which
sequences in the corpus appear next to each other. For pretraining transformer-based neural language models for the aviation domain (with orthographic
transcriptions), this same procedure can be used to classify whether two transmissions occur in the correct order. Generalizing a bit, we can consider
two individual transmissions to be $A$ and $B$; the training objective for the network is then to predict whether transmission $B$ occurs in response to
transmission $A$. This would be considered a binary classification task; $B$ either occurs in response to $A$ or it does not. In practice, for a true
label (meaning $B$ does, in fact, occur in response to $A$), transmission $A$, for example, can be a pilot request to ATC and transmission $B$ would
be ATC's response. A false label (meaning $B$ does not occur in response to $A$) would be made up of transmissions randomly shuffled such that $A$ and
$B$ do not occur together in the corpus. Note that, since these are only two transmissions, neither $A$ nor $B$ will necessarily be one side of the
transmission i.e.~$A$ can be either a pilot or ATC. Of the four corpora used in this work, ATCC is the only corpus that can be mined for an NSP-like
pretraining procedure since the raw transcripts have indicators for time (start and end indicators), transmitting party, and intended receiver. With
this information, we can estimate the correct order of transmissions between two parties to a reasonable degree of certainty (manually or
programmatically). For example, using the following transmissions from ATCC corpus, in which the Boston-Logan tower is communicating with PAA540. The
transmission below can be used as $A$:

\begin{quote}
    CLIPPER SIXTY FIVE FORTY IS AH FOUR FROM RIPIT CLEARED FOR THE I L S D M E TWO SEVEN APPROACH
\end{quote}

\noindent
and the response from PAA540 to the Boston-Logan tower can be used as $B$:

\begin{quote}
    AH ROGER CLEARED FOR THE APPROACH AH CLIPPER FIVE FORTY THANK YOU
\end{quote}

\noindent
In this case, $A$ and $B$ occur in the middle of an interaction. There are two transmissions between PAA540 and the Boston-Logan tower that occur prior
to the two above. We can consider the two instances of $A$ and $B$ above as a true label, meaning they occur one after the other in the corpus. To
construct a false label, we can simply reverse the input of $A$ and $B$, i.e.~$A$ becomes $B$ and $B$ becomes $A$, or choose another transmission at
random from the corpus to replace either $A$ or $B$. For proper construction of the corpus for the NSP pretraining task, the distribution of true and
false labels should be evenly balanced. If there are 500 samples labeled as true, then there should be 500 samples labeled as false.
